---
title             : "Plant life history strategies predicted by satellite-detected drought"
shorttitle        : ""

author: 
  - name          : "J Grey Author"
    affiliation   : "1,2"
    corresponding : yes    # Define only one corresponding author
    address       : "307 University Ave, Fort Collins, CO 80523"
    email         : "monroejg@colostate.edu"
  - name          : "Brian Gill"
    affiliation   : "3"
  - name          : "Brian Gill"
    affiliation   : "4"
    
affiliation:
  - id            : "1"
    institution   : "Graduate Degree Program in Ecology, Colorado State University"
  - id            : "2"
    institution   : "College of Agriculture, Colorado State University"
  - id            : "3"
    institution   : "Department of Ecology and Evolutionary Biology"
  - id            : "4"
    institution   : "Biology Department, Penn State University"    
    

authornote: |
  Add complete departmental affiliations for each author here. Each new line herein must be indented, like this line.

abstract: |

  <!-- https://tinyurl.com/ybremelq -->
  
keywords          : "drought, life history, remote sensing, phylogeography"
wordcount         : ""

bibliography      : ["r-references.bib"]

floatsintext      : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no
figsintext        : yes

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
---

```{r setup, include = FALSE}
library("papaja")
library(raster)
library(tidyverse)
library(ggplot2)
```

#Introduction

Life history 
Annual vs perennial
Hypotheses
Summary


# Methods
## Data
### Life history
### Phyologeny
### Herbarium Specimens
```{r raw_GBIF}
GBIF<-read.csv("../data/0006178-160311141623029/occurrence.csv")
```
We downloaded `r nrow(GBIF)` records from GBIF. 

### Drought

## Analyses
### Phylogeny
### Drought frequency
### Contrast
We used `r cite_r("r-references.bib")` for all our analyses.

# Results

(ref:foo) A scatterplot of the data `cars` using **base** R graphics. 

```{r foo, fig.cap='(ref:foo)',out.width = "\\textwidth", fig.pos = "!h"}
plot(cars)  # a scatterplot
```

First we did this (Fig. \@ref(fig:foo).)

# Discussion
The best paper ever [@monroe2016adaptation]

There is a package installer at pandoc’s download page. If you later want to uninstall the package, you can do so by downloading this script and running it with perl uninstall-pandoc.pl.There is a package installer at pandoc’s download page. If you later want to uninstall the package, you can do so by downloading this script and running it with perl uninstall-pandoc.pl.There is a package installer at pandoc’s download page. If you later want to uninstall the package, you can do so by downloading this script and running it with perl uninstall-pandoc.pl.There is a package installer at pandoc’s download page. If you later want to uninstall the package, you can do so by downloading this script and running it with perl uninstall-pandoc.pl.There is a package installer at pandoc’s download page. If you later want to uninstall the package, you can do so by downloading this script and running it with perl uninstall-pandoc.pl.There is a package installer at pandoc’s download page. If you later want to uninstall the package, you can do so by downloading this script and running it with perl uninstall-pandoc.pl.


\newpage

# References
```{r create_r-references}
r_refs(file = "r-references.bib")
```

\begingroup
\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}

<div id = "refs"></div>
\endgroup
